---
title: "Hands on Exercise 2"
author: "Dhreeti"
---

# 2.1 Overview

Introduction to several ggplot2 extensions for creating more elegant and effective statistical graphics

## 2.2 Getting Started

First we will check and load our packages

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 

```

Importing Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")

```

## 2.3 Plotting with ggrepel

The issue with this is regarding the annotation which makes it hard to see.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

```

### 2.3.1 Working with ggrepel

This is plot of english vs math scores for Primary 3 students

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

```

## 2.4 ggplot2 Themes

Showcasing graph with themes

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 


```

### 2.4.1 Working with ggtheme package

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

### 2.4.2 Working with hrbthems packag

```{r warning: false}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()


```

### 2.4.3 A more refined graph

```{r warning: false}

ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")


```

# 2.5 Beyond Single Graph

Graph 1

```{r}
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")


```

Graph 2

```{r}
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")


```

Graph 3 - Scatter plot

```{r}
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")


```

### 2.5.1 Creating Composite Graphics: pathwork methods

Combining 2 graphs using P1 + P2

```{r}
p1 + p2

```

Combining 3 graphs

```{r}
(p1 / p2) | p3

```

### 2.5.2 Creating a composite figure with tag

```{r}

((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')

```

### 2.5.3 Creating figure with insert

```{r}
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)

```

### 2.5.4 Creating a composite figure by using patchwork and ggtheme

```{r}
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

## 2.6 Extra plotting to practice 

Gender Represenation By Class

```{r}
ggplot(exam_data, aes(x = CLASS, fill = GENDER)) +
  geom_bar(position = "fill") +
  labs(title = "Gender Representation by Class", y = "Proportion", x = "Class") +
  theme_tufte() +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"))



```

 Subject Score Distribution by Race


```{r}
ggplot(exam_data, aes(x = RACE, y = MATHS, fill = RACE)) +
  geom_violin() +
  labs(title = "Math Scores by Race", y = "Math Scores", x = "Race") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
