---
title: "Hands on Exercise 5d"
author: "Dhreeti"
execute: 
  echo: true
  warning: false
---

# 1.Overview

hands-on experience on:

plotting statistic parallel coordinates plots by using ggparcoord() of GGally package, plotting interactive parallel coordinates plots by using parcoords package, and plotting interactive parallel coordinates plots by using parallelPlot package.

# 2. Installing and Launching R Packages

```{r}

pacman::p_load(GGally, parallelPlot, tidyverse)

```

# 3 Data Preparation

```{r}

wh <- read_csv("data/WHData-2018.csv")

```

# 4 Plotting Static Parallel Coordinates Plot

## 4.1 Plotting a simple parallel coordinates

Code chunk below shows a typical syntax used to plot a basic static parallel coordinates plot by using ggparcoord().

```{r}


ggparcoord(data = wh, 
           columns = c(7:12))

```

## 4.2 Plotting a parallel coordinates with boxplot

```{r}

ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")


```

## 4.3 Parallel coordinates with facet

Since ggparcoord() is developed by extending ggplot2 package, we can combination use some of the ggplot2 function when plotting a parallel coordinates plot.

In the code chunk below, facet_wrap() of ggplot2 is used to plot 10 small multiple parallel coordinates plots. Each plot represent one geographical region such as East Asia.

```{r}

ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)


```

## 4.4 Rotating x-axis text label

```{r}


ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))


```

## 4.5 Adjusting the rotated x-axis text label

```{r}

ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))


```

# 5.Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

## 5.1 The basic plot

```{r}

wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)


```

## 5.2 Rotate axis label

```{r}

parallelPlot(wh,
             rotateTitle = TRUE)

```

## 5.3 Changing the colour scheme

```{r}

parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)

```

## 5.4 Parallel coordinates plot with histogram

```{r}

histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)


```
